Drought vulnerability of tree growth

### 2024-06-27 scripts/explore_analysis.Rmd
data summaries; resistance for 2010 and 2014

---
title: "Climate Sensitivity of Tropical Tree Growth"
author: ""
output: html_document
---

@@

#this can be refined if needed
wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)

@@

census_int<-dendrobyband%>%
  group_by(Cno) %>%
  dplyr::summarise(start=min(Date2, na.rm = T),
                   end=max(Date2, na.rm=T))%>%
  dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))

pre_yr_census<-ggplot()+
  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
  # scale_y_continuous(    # Features of the first axis
  #   name = "monthly rainfall (mm)",
  #   
  #   # Add a second axis and specify its features
  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("Date")+theme_bw()+
  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

pre_yr_census

@@

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  filter(Cno %in% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(year, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))

@@

# species_all<-dendro.data.noouts %>%
#   group_by(sp, treeID) %>%
#   dplyr::summarise(n.obs=length(treeID))%>%
#   ungroup() %>%
#   group_by(sp) %>%
#   dplyr::summarise(n.tree=length(treeID),
#                    n.obs.tot=sum(n.obs))%>%
#   ungroup()%>%
#   arrange(-n.tree)
# 
# write.csv(species_all, "../data/HKK-dendro/species.csv", row.names = F)

@@

```{r echo=FALSE, eval=F}

@@

## Detrending all species

growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend_all

```
## Fastest and slowest in 2015 and 2016

@@

growth_2014 <- dendro.data.noouts %>%
  filter(year==2014)

@@

png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )

@@

>>>>

---
title: "Drought vulnerability of tree growth"
author: ""
output: html_document
---

@@

#this can be refined if needed
#wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)

@@

census_int<-dendrobyband%>%
  group_by(Cno) %>%
  dplyr::summarise(start=as.Date(min(Date2, na.rm = T)),
                   end=as.Date(max(Date2, na.rm=T)))#%>%
  #dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))

pre_yr_census<-ggplot()+
  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_rect(data=census_int, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
              fill = "grey10", alpha = 0.5) +
  geom_text(data=census_int, aes(x=start, y=500, 
                                 label= paste0("C", 1:29), angle=90), size=2)+
  # scale_y_continuous(    # Features of the first axis
  #   name = "monthly rainfall (mm)",
  #   
  #   # Add a second axis and specify its features
  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("")+theme_bw()+ ylab("total monthly precipitation (mm)")+
  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

png("../doc/display/census_precip.png", width=6, height=3, res=300, units="in") 
pre_yr_census
dev.off()

@@

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  #filter(Cno %in% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(Cno, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~Cno, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

@@

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = Cno, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  filter(abs(std.res)<=3)%>%
  select(-c(std.res, out))
  #mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) #%>% #create an outlier removed column

```
```{r calculating increments for specific years}

dendro_2010<- dendro.data.noouts %>%
  filter(Cno %in% c(1, 3, 5))%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year(Date2)-lag(year(Date2)),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  dplyr::mutate(
    diainc.scaled = scale(inc_annual),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))#%>%
  #filter(out.inc==0)

quantiles_2009<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==3], 
                         probs=c(0.25, 0.5, 0.75, 1))

quantiles_2010<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==5], 
                         probs=c(0.25, 0.5, 0.75, 1))

dendro_2010_plot<-ggplot(data = dendro_2010, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
  geom_density(alpha=0.5) +
  scale_color_manual(values=c("forestgreen", "brown"))+
  scale_fill_manual(values=c("forestgreen", "brown"))+
  geom_vline(xintercept=quantiles_2009, col="forestgreen")+
  geom_text(aes(x=0.75, y=1, label="2009"), col="forestgreen", size=3)+
  geom_text(aes(x=0.75, y=1.5, label="2010"), col="brown", size=3)+
  geom_vline(xintercept=quantiles_2010, col="brown")+ 
  xlab("diameter increment (cm)") + ylab("density")+
  guides(col="none", fill="none")+
  theme_bw()

dendro_2015<- dendro.data.noouts %>%
  filter(Cno %in% c(11, 13, 15))%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year(Date2)-lag(year(Date2)),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  dplyr::mutate(
    diainc.scaled = scale(inc_annual),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

quantiles_2014<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==13], 
                         probs=c(0.25, 0.5, 0.75, 1))

quantiles_2015<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==15], 
                         probs=c(0.25, 0.5, 0.75, 1))

dendro_2015_plot<-ggplot(data = dendro_2015, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
  geom_density(alpha=0.5) +
  scale_color_manual(values=c("forestgreen", "brown"))+
  scale_fill_manual(values=c("forestgreen", "brown"))+
  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
  geom_vline(xintercept=quantiles_2015, col="brown")+ 
  xlab("diameter increment (cm)") + ylab("density")+
  guides(col="none", fill="none")+
  theme_bw()

library(gridExtra)
png("../doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
grid.arrange(dendro_2010_plot, dendro_2015_plot, nrow=1)
dev.off()

```
```{r drought resistance}

dec_williams<-read.csv("../data/HKK-dendro/species.csv")

dec_williams<-dec_williams %>%
  filter(n.tree >= 5)

resistance_2010 <- dendro_2010 %>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
         resistance.div = inc_annual/inc_min1,
         resistance.dif = inc_annual-inc_min1)%>%
  filter(!is.na(inc_min1), !is.infinite(resistance.div))

resistance_2010<-merge(resistance_2010, dec_williams, by="sp", all.x=T)

resistance_2010_div<-ggplot(data = resistance_2010, aes(x=resistance.div)) +
  geom_density(alpha=0.5) +
  theme_bw()

resistance_2010_dif<-ggplot(data = resistance_2010, aes(x=resistance.dif)) +
  geom_density(alpha=0.5) +
  theme_bw()

resistance_2015 <- dendro_2015 %>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
         resistance.div = inc_annual/inc_min1,
         resistance.dif = inc_annual-inc_min1)%>%
  filter(!is.na(inc_min1), !is.infinite(resistance.div))

resistance_2015<-merge(resistance_2015, dec_williams, by="sp", all.x=T)

resistance_2015_div<-ggplot(data = resistance_2015, aes(x=resistance.div)) +
  geom_density(alpha=0.5) +
  theme_bw()

resistance_2015_dif<-ggplot(data = resistance_2015, aes(x=resistance.dif)) +
  geom_density(alpha=0.5) +
  theme_bw()

png("../doc/display/resistance_dif.png", width=8, height=4, units="in", res=300)
grid.arrange(resistance_2010_dif, resistance_2015_dif, nrow=1)
dev.off()

```
```{r basic model}

model_2010<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
                  calcDBH_min1*cii_min1 + 
                  calcDBH_min1*williams_dec+
                  cii_min1*williams_dec,
                data=resistance_2010)

summary(model_2010)

model_2015<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
                  calcDBH_min1*cii_min1 + 
                  calcDBH_min1*williams_dec+
                  cii_min1*williams_dec,
                data=resistance_2015)

summary(model_2015)

resistance_all<-rbind(resistance_2010, resistance_2015)
resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)))

resistance_model<-resistance_all %>%
  select(resistance.dif, calcDBH_min1, cii_min1, williams_dec, yr)

resistance_model<-resistance_model[complete.cases(resistance_model),]

library(lme4)
library(nlme)
model_all<-lme(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
                  calcDBH_min1*cii_min1 + 
                  calcDBH_min1*williams_dec+
                  cii_min1*williams_dec,
                random=~1|yr,
                data=resistance_model)

model_all
anova(model_all)

library(nlme)
library(ggplot2)

# Create a new data frame for predictions
new_data <- expand.grid(
  calcDBH_min1 = seq(min(resistance_model$calcDBH_min1), max(resistance_model$calcDBH_min1), length.out = 100),
  cii_min1 = seq(min(resistance_model$cii_min1), max(resistance_model$cii_min1), length.out = 100),
  williams_dec = seq(min(resistance_model$williams_dec), max(resistance_model$williams_dec), length.out = 100)
)

# Generate predictions
new_data$predicted <- predict(model_all, newdata = new_data, level = 0)

# Create the plot
ggplot(resistance_model, aes(x = williams_dec, y = resistance.dif)) +
  geom_point() +
  geom_line(data = new_data[new_data$cii_min1==2,], aes(y = predicted)) +
  theme_minimal() +
  labs(x = "Deciduousness", y = "Drought resistance", title = "Predictions from LME Model")

@@

@@

## Fastest and slowest in 2015 and 2016

@@

growth_2014 <- dendro.data.noouts %>%
  filter(Cno==11)

@@

png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
dev.off()

@@

dec_williams<-read.csv("../data/HKK-dendro/species.csv")

dec_williams<-dec_williams %>%
  filter(n.tree >= 5)

trial<-merge(growth_2014, dec_williams, by="sp")

png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
ggplot(trial, aes(x = williams_dec, y = factor(Cii), 
                       fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
  labs(title = 'deciduousness', y="canopy illumination") +
  #theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8)
    )
dev.off()
